---
title: "Classifying e-obs acceleration & GPS data into flying/not-flying"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: cerulean
    highlight: zenburn
---



# Set workspace

```{r global_options, include=FALSE}

# Global options
# fig_path = "Rmd_output"
# if( !dir.exists(fig_path)) dir.create(fig.path)

knitr::opts_chunk$set(fig.width=12, fig.height=8,
                      # fig.path=fig_path,
                      echo=TRUE, warning=FALSE, message=FALSE)

options(DT.options = list(pageLength = 25,
                          dom = 'Btip',
                          filter = "top",
                          extensions = 'Buttons',
                          buttons = c('copy', 'csv')))


```


```{r}

# ==============================================================================
# Work with gps data downloaded from movebank
# ==============================================================================

# Download method:
# ---------------
# Go to: https://www.movebank.org/panel_embedded_movebank_webapp
# Click on Studies on the menu bar on the left side of the viewer
# Choose my study
# Click on Download data
# Choose: csv, include undeployed locations,
#         add UTM coordinates, add study local time
#         (see screenshot)
# Download above for both gps and acceleration data
#
# NOTE: Don't open downloaded files in excel, because it's somehow corrupting
#       the date column, making it difficult for R to read it correctly



# ==============================================================================
# * Environment
# ==============================================================================
source("R/00_fxns.R")

library(maptools)
library(move)
library(dplyr)
library(data.table)

```

# Import acceleration and GPS data

```{r import-data}

ACC_file = file.path("data/raw/e-obs", "movebank_acc.csv") # acc file
GPS_file = file.path("data/raw/e-obs", "movebank_gps.csv") # gps file
crop_file = file.path("data/raw/bats", "eobs_info.csv") # time-crop file

out_dir = "data/acc_behavL/"
if( !dir.exists(out_dir)) dir.create(out_dir)

out_filename = "acc_gps_behavL.RDS" # the axis info is appended to this filename

axises = c("xy", "xz", "yz", "xyz") # choose one or as many options as you want

# ==============================================================================
# * Read in acceleration and GPS data
# ==============================================================================

# read in the acceleration data
accALL = read.csv(ACC_file, stringsAsFactors = FALSE)

# read in the gps data and assign local timestamp as main one
gpsALL = read.csv(GPS_file, stringsAsFactors = FALSE)

# read in csv with timestamp-crop data
time_crop = read.csv(crop_file, stringsAsFactors = FALSE)

```

# Prepare acceleration and GPS data 

```{r, eval = TRUE, results = 'hide'}

# ==============================================================================
# * Get acceleration and GPS data ready for processing
# ==============================================================================

# ** crop ACC anf GPS data to retain bat-specific data
# ------------------------------------------------------------------------------
# This also removes data for tags 5314 & 5326, which had very sparse data
tags = unique(time_crop$tagID)

accALL = lapply(tags, function(ID, time_crop, accALL) {
    df = accALL[ accALL$tag.local.identifier %in% ID, ]
    start_ts = time_crop$start_ts_UTC[ time_crop$tagID %in% ID ]
    stop_ts = time_crop$stop_ts_UTC[ time_crop$tagID %in% ID ]
    df[ df$timestamp >= start_ts &
        df$timestamp < stop_ts, ]
}, time_crop, accALL) %>%
    rbindlist


gpsALL = lapply(tags, function(ID, time_crop, gpsALL) {
    df = gpsALL[ gpsALL$tag.local.identifier %in% ID, ]
    start_ts = time_crop$start_ts_UTC[ time_crop$tagID %in% ID ]
    stop_ts = time_crop$stop_ts_UTC[ time_crop$tagID %in% ID ]
    df[ df$timestamp >= start_ts &
        df$timestamp < stop_ts, ]
}, time_crop, gpsALL) %>%
    rbindlist



# ** switch timestamp to local time and backup original to eobs.timestamp
# ------------------------------------------------------------------------------
accALL$eobs.timestamp = accALL$timestamp
accALL$timestamp = accALL$study.local.timestamp

gpsALL$eobs.timestamp = gpsALL$timestamp
gpsALL$timestamp = gpsALL$study.local.timestamp

# tell R what the local timestamp format is
accALL$timestamp = as.POSIXct(accALL$timestamp, format="%Y-%m-%d %H:%M:%OS",
                              tz="Africa/Dar_es_Salaam")

gpsALL$timestamp = as.POSIXct(gpsALL$timestamp, format="%Y-%m-%d %H:%M:%OS",
                              tz="Africa/Dar_es_Salaam")


# ** populate the individual local identifier column with tag ID info
# ------------------------------------------------------------------------------
# (needed by the move package)
accALL$tagID = paste0("K", accALL$tag.local.identifier)
gpsALL$tagID = paste0("K", gpsALL$tag.local.identifier)


# ** for GPS data, keep those with status "A", which is the good quality data
# ------------------------------------------------------------------------------
gpsALL = gpsALL[ gpsALL$eobs.status %in% "A", ]


# ** make a list of acceleration & gps data
# ------------------------------------------------------------------------------
accL = split(accALL, accALL$tagID)
gpsL = split(gpsALL, gpsALL$tagID)

# check if individuals names are the same in acc file as in the move object
identical(sort(names(gpsL)), sort(names(accL))) 

```
	
# Classify acceleration & GPS data to flying or not-flying

## Acc burst classification rationale
Acceleration burst data is one long string of acc recordings along 3-axes: X, Y, and Z.
With an ACC BYTE COUNT of 1135 bytes, the each ACC burst corresponds to 792 sequential recordings along the X, Y, and Z axes. 

E.g. a reading of an acc burst:

```{r acc-explanation, eval = TRUE}

accL[[1]]$eobs.accelerations.raw[1]

```

Going forward, with each acc burst we extract individual readings using `strsplit` by `space`, and create a matrix of the transposed acc readings such that each row corresponds to an individual acc burst, and each column represents the sample number within the burst, for a total of 792 columns. We then extract every first, second, and third column to get subsets of data corresponding to only X, Y, or Z readings. 


### E.g. using hypothetical data with 9 readings for each burst

```{r acc-eg, eval = TRUE}

acc_gen = function() paste0(sample(1:200, 9), c("x", "y", "z"), collapse = " ")
    
acc_df = data.frame(acc = c(acc_gen(), acc_gen(), acc_gen()),
                stringsAsFactors = FALSE)

acc_df

acc_df2 = strsplit(acc_df$acc, split = " ", fixed = TRUE) 
acc_matrix = matrix(unlist(acc_df2), ncol = 9, byrow = TRUE) 
colnames(acc_matrix) = paste0("V", 1:9)

acc_matrix

# X axis readings
acc_matrix[, seq(1, 9, 3)]

# Y axis readings
acc_matrix[, seq(2, 9, 3)]

# Z axis readings
acc_matrix[, seq(3, 9, 3)]

```

Once we subset the acc readings by each axis, we can calculate the variance of each burst's readings and do a k-cluster analyses to classify acc bursts by activity (flying vs. not flying).


## Check lengths of acc bursts
	
Based on the tag settings, our ACC BYTE COUNT was 1135 bytes, i.e. 792 ACC samples would be collected at each burst. 

```{r, eval = TRUE}

# Determining lengths of samples per acc burst
accaxesL = sapply(accL, function(acc) {
    accaxes = strsplit(as.character(acc$eobs.accelerations.raw),
                       split=" ", fixed=T) 
    paste0(sort(unique(sapply(accaxes, length))), collapse = ", ")
})

data.frame(tagID = names(accaxesL), numberSamples = accaxesL) %>%
    mutate(row.names = NULL)

```


## Classification process

The functions used below have been built upon code by Abedi-Lartey et al.:

Abedi-Lartey M, Dechmann DKN, Wikelski M, Scharf AK, Fahr J. Long-distance seed dispersal by straw-coloured fruit bats varies by season and landscape. Global Ecology and Conservation. 2016;7:12–24. Available from: https://doi.org/10.1016/j.gecco.2016.03.005

```{r acc-classify, eval = TRUE, results = 'hide'}

numberSamples = 792

for( axis in axises) {
    acc_activityL = lapply(accL, get_acc_activity, numberSamples = 792, axis = axis)
    tagIDs = names(acc_activityL)
    acc_gps_behavL = lapply(tagIDs, get_acc_gps_behav, acc_activityL, gpsL)
    names(acc_gps_behavL) = tagIDs

    # save data
    saveRDS(acc_gps_behavL, file.path(out_dir, paste0(axis, "_", out_filename)))
}


```

# Session info

```{r}
sessionInfo()
```
